Forecasting the 2024 U.S. Presidential Election: A Poll-of-Polls Approach for Predicting the Outcome

### 2024-11-03 paper/paper.qmd
3

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - First author
  - Another author
thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(palmerpenguins)
```

@@

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

@@

We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...

Overview text

@@

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

@@

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

@@

In this study, we employed a Generalized Linear Model (GLM) to predict the outcome of the upcoming US presidential election. Specifically, we used polling data to forecast the vote percentage for each candidate. The model was developed using the poll-of-polls methodology, which aggregates data across multiple polls, allowing for a more robust prediction by reducing individual poll biases. [Appendix -@sec-model-details].

@@

>>>>

---
title: "Forecasting the 2024 U.S. Presidential Election: A Poll-of-Polls Approach for Predicting the Outcome"
subtitle: "Applying Aggregated Poll Data and Statistical Model to Reveal a Narrow Path to Victory in a Highly Competitive Race"
author: 
  - Chendong Fei
  - Xinze Wu
  - Claire Ma
thanks: "Code and data are available at: https://github.com/ke3w/Prediction_US_presidential_election.git"
date: today
date-format: long
abstract: "This paper develops a model to forecast the outcome of the 2024 U.S. presidential election by analyzing aggregated polling data, or “poll-of-polls,” sourced from FiveThirtyEight. Using a generalized linear model, we assess national trends alongside key battleground state polls to predict each candidate's likelihood of victory. The findings indicate a closely contested race, with specific demographic and regional factors creating narrow pathways to winning the presidency. This analysis highlights the value of aggregated polling data in understanding electoral dynamics and demonstrates the importance of statistical modeling in making informed predictions about major political events."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(ggplot2)
library(MASS)
library(dplyr)
```

@@

The outcome of the U.S. presidential election has far-reaching implications, shaping both domestic policies and international relations. As the 2024 election approaches, voters and analysts turn to polls to understand the state of the race between Vice President Kamala Harris, the Democratic candidate, and former President Donald Trump, the Republican candidate. However, individual polls are often limited by their methodologies, timing, and sample demographics, leading to variations in predictions. To overcome these limitations, aggregating multiple polls—a technique known as “poll-of-polls”—provides a more stable and reliable indicator of public opinion. This paper applies a poll-of-polls approach, informed by methodologies from individual polls[@blumenthal2014;@pasek2015], to predict the outcome of the 2024 U.S. presidential election, focusing on data aggregated by FiveThirtyEight[@fivethirtyeight2024].

The primary objective of this analysis is to forecast which candidate is likely to win the 2024 election based on aggregated national and battleground state polling data. By constructing a generalized linear model, we aim to distill insights from the extensive polling data available, examining trends and key demographic indicators.

The primary estimand in this analysis is the probability of each candidate winning the 2024 U.S. presidential election based on aggregated polling data. This probability is derived from a weighted average of poll results across national and battleground states, with adjustments for factors such as recent polling trends, sample sizes, and state-specific electoral significance.

Our analysis reveals a highly competitive race, with key battleground states playing a pivotal role in determining the overall outcome. The model identifies specific regions and demographics that are likely to influence the election results, highlighting the polarized nature of the electorate.As of November 1, 2024, FiveThirtyEight’s national polling average indicates a slight edge for Harris, who has 48.1% support compared to Trump’s 46.7%. Despite this narrow national lead, the race in critical battleground states remains highly competitive. For instance, Pennsylvania is evenly split, with Harris holding marginal leads in states like Wisconsin and Michigan, while Trump shows slight advantages in Nevada, Georgia, and Arizona. These tight margins highlight the crucial role battleground states play in determining the election’s outcome.

These findings underscore the importance of aggregated poll data in capturing the broader political landscape, offering insights that single polls may miss. By understanding the dynamics at play, this study contributes to a broader understanding of electoral processes and the predictive power of statistical models in forecasting complex political events.

This paper is organized as follows: @sec-data discusses the details of the dataset. @sec-Methodology describes the methodology, including generalized linear model. @sec-Results presents the results, highlighting trends in polling, and @sec-Discussion considers the implications of these results for future research on polling and public opinion.

@@

We use the statistical programming language R [@citeR] to analyze polling data from FiveThirtyEight’s U.S. Presidential election polls [@fivethirtyeight2024]. The dataset contains  information such as pollster, polling date, methodology, sample size, state, and candidate support percentages. It allows us to track voter sentiment across different regions and polling methods, providing a comprehensive view of the election landscape. Additionally, irrelevant or incomplete entries were removed to ensure clean, high-quality data, and we retained only key variables to streamline the analysis. This careful selection and cleaning process ensure that the dataset offers a precise and representative snapshot of the election landscape.

@@

The dataset measures public opinion on the 2024 U.S. presidential election by aggregating polling data to estimate voter support for each candidate at both national and state levels. These polling data entries are then aggregated, which applies a weighted adjustment to reflect the reliability, sample size, and recency of each poll. This weighting process addresses the natural variation in polling methodologies (e.g., online survey, phone), sample diversity, and timing, which influence the reliability of each poll as a measure of the broader population’s preferences.For instance, if this poll was conducted a week before the election, it might be weighted more heavily than a poll from three months prior, as it better represents current voter sentiment. By weighting higher-quality and more recent polls more heavily, it creates a comprehensive measure that accounts for both regional and national voter sentiment, smoothing out biases from individual polls.

@@

In our analysis, the primary outcome variable is labeled `win`, which is a binary indicator representing the likelihood of a candidate "winning" in each poll based on their support percentage. Specifically, `win` is defined as follows: if a candidate's support percentage (`pct`) in a given poll exceeds 50%, then `win` is assigned a value of 1, indicating a projected win for that candidate in that poll. If the support percentage is 50% or below, `win` is assigned a value of 0, indicating that the candidate is not the likely winner in that poll.
This binary outcome variable is particularly useful for logistic regression analysis, as it allows us to model the probability of a candidate achieving majority support in each poll. Using win provides a clear and interpretable framework to assess factors influencing a candidate’s chances of gaining majority support, which aligns well with election forecasting goals. Additionally, this threshold reflects the electoral concept of a "win," as it represents the point at which a candidate has more than half of the vote share, an essential consideration in political analysis.

## Predictor variables
The predictor variables in this analysis were chosen based on their potential influence on polling outcomes and candidate support. Each predictor reflects characteristics of the poll, the pollster, or the candidate's support environment. These variables aim to capture the factors that could impact the likelihood of a candidate reaching majority support (win = 1). Key predictor variables include:

- **sample_size**: Represents the number of respondents in each poll, with larger sample sizes generally leading to more reliable results.
- **pollster Rating**: Indicates the quality and historical accuracy of the polling organization, helping to account for differences in poll reliability.
- **state**: Captures the geographical region of the poll, reflecting regional differences in voter support that are crucial in U.S. elections.

These three variables provide a balanced view of poll quality, reliability, and regional influence, enhancing the model's ability to predict election outcomes accurately. This combination ensures that the model is interpretable and captures essential factors influencing voter sentiment. We apply Bayesian Information Criterion (BIC) method to select significant predictors, and a summary statistics for this method shown in @Appendix

```{r}
# Define the binary response variable if not already defined
# Create a binary `win` variable where pct > 50% indicates a "win"
data <- data %>%
  mutate(win = ifelse(pct > 50, 1, 0))
# Fit the full logistic regression model with all predictors
full_model <- glm(
  win ~ sample_size + pollscore + transparency_score + 
          methodology + start_date + end_date + state + party, 
  data = data, 
  family = binomial
)
# Use stepwise selection based on BIC
best_model <- stepAIC(
  full_model, 
  direction = "both", 
  k = log(nrow(data)) # Set k to log(n) for BIC
)

# Display the summary of the selected model based on BIC
summary(best_model)
# View final model formula and BIC value
formula(best_model)
BIC(best_model)
```

@@

In this study, we employed a Generalized Linear Model (GLM) to predict the outcome of the upcoming US presidential election. Specifically, we used polling data to forecast the vote percentage for each candidate. The model was developed using the poll-of-polls methodology, which aggregates data across multiple polls, allowing for a more robust prediction by reducing individual poll biases.

@@

##Exploratory Data Analysis (EDA) Plots

As shown in Figure @fig-vote-distribution, the distribution of vote percentages varies by candidate, with some candidates demonstrating stronger support than others.
```{r}
#| label: fig-vote-distribution
#| fig-cap: "Distribution of Vote Percentage by Candidate"
#| echo: false

analysis_data %>%
  ggplot(aes(x = pct, fill = candidate_name)) +
  geom_histogram(binwidth = 5, alpha = 0.7) +
  labs(title = "Distribution of Vote Percentage by Candidate",
       x = "Vote Percentage",
       y = "Count",
       fill = "Candidate") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8),
        legend.text = element_text(size = 6))
```

Average Poll Score by State: Figure @fig-avg-pollscore shows the average poll score by state, providing insight into regional differences in polling support. This visualization helps identify states where particular candidates may have stronger or weaker support.
```{r}
#| label: fig-avg-pollscore
#| fig-cap: "Average Poll Score by State"
#| echo: false

analysis_data %>%
  group_by(state) %>%
  summarise(avg_pollscore = mean(pollscore, na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(state, -avg_pollscore), y = avg_pollscore)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Average Poll Score by State",
       x = "State",
       y = "Average Poll Score") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 5)) +
  coord_flip()
```

Overall, the diagnostic evaluation suggested that the model provided a good fit for the data, with no major violations of GLM assumptions.
